---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# sfcm

<!-- badges: start -->
[![Lifecycle: experimental](https://img.shields.io/badge/lifecycle-experimental-orange.svg)](https://lifecycle.r-lib.org/articles/stages.html#experimental)
<!-- badges: end -->

The goal of **sfcm** is to provide tools for estimating bird collisions in planned wind farms using a stochastic
Flux Collision Model, which allows for modelling uncertainty in parameter estimates via simulation. 

## Features

- Incorporate uncertainty in parameter estimates in bird collision estimates.
- Simulate parameters from multiple probability distributions (normal, Poisson, negative binomial and beta).
- Lightweight and fast.

## Installation

You can install the development version of sfcm from [GitHub](https://github.com/jvdz-aw/sfcm) with:

```{r eval = FALSE}
# install.packages("pak")
pak::pak("jvdz-aw/sfcm")
```

Or using:

```{r eval = FALSE}
# install.packages("devtools")
devtools::install_github("jvdz-aw/sfcm")
```

Alternatively, you can download the binary version you prefer and install it with:

```{r eval = FALSE}
install.packages("sfcm_0.1.0.tar.gz")
```

## Example

This is an example of the basic workflow, which shows you how to create a test dataset, perform a simulation
and estimate the number of bird collisions:

```{r example, eval = FALSE}
library(sfcm)

# Create a test dataset
test_df <- data.frame(
  species = c("A", "B"),
  flux_mean = c(50, 100),
  flux_sd = c(10, 50),
  a_macro = 0.95,
  f_prop = 1,
  h_prop = 0.6,
  h_prop_ref = 0.67,
  rotor_d = 170,
  rotor_d_ref = 60,
  turb_dist = 600,
  turb_dist_ref = 250,
  turbs_e_mean = c(4, 2),
  turbs_e_sd = 0,
  turbs_e_ref = 4,
  p_col_mean = c(0.5, 0.8),
  p_col_sd = c(0.05, 0.1)
)

# Define what parameters to simulate and which probability distributions to use for random sampling
pars <- c("flux", "turbs_e", "p_col")
dists <- c("nbinom", "poisson", "beta")

# Perform simulation using n = 10 iterations
sim_out <- simulate_parameters(test_df, parameters = pars, distributions = dists, n = 10)

# The simulation output is a `model_input` S3 class object, which we use for implicit data validation
class(sim_out)

# Estimate bird collision using the wrapper function
res <- run_model(sim_out)
```

## Supported parameters and distributions

The following table shows which probability distributions are supported for each parameter that can be simulated.

| Parameter | Supported probability distribution |
|-----------|------------------------------------|
| flux      | Normal, Poisson, negative binomial |
| turbs_e   | Poisson                            |
| p_col     | Beta                               |

## Contributing

Please open issues or submit pull requests on [GitHub](https://github.com/jvdz-aw/sfcm)

## License

This package is licensed under the MIT license. See the LICENSE file for details.

## Acknowledgements

The development of this R package is a collaboration between [Altenburg & Wymenga](https://www.altwym.nl/) and [Waardenburg Ecology](https://waardenburg.eco/).
